The Effects of the NBA COVID Bubble on the NBA Playoffs: A Case Study for Home-Court Advantage

### 2022-02-12 Manuscript/manuscript.tex
Update manuscript.tex
aded reference to Bonferroni correction
Update manuscript.tex
Added study about referee bias in NBA

There are also surveys on the factors contributing to home-court advantage.
\citet{Carron1992} gave four main game location factors for home and away
teams, namely, the crowd factor, which is the impact of fans cheering; learning
factors, which is an advantage from home teams from playing at
a familiar venue; travel factors, the idea that away teams may face
fatigue and jet lag from traveling; and, rule factors, which says that home teams
may benefit from some advantages in rules and officiating. They acknowledge that
these factors would all be removed if games were played at a neutral site, even
if one team was designated as ``home team''. This study was reviewed a decade later
by \citet{Carron2005}. The 2005 review goes over the new findings from studies
about the significance of these four game location factors. Since 1992, they have found
that results on these four factors are mixed. However, there is some evidence
supporting crowd and travel factors impact games across all major sports. There is less
evidence suggesting learning and rule factors impact across the various collegiate 
and professional sports. The NBA is not a league which has rules that
may favor the home team, like batting last in the MLB, but these rule
factors also account for referee bias which may impact the NBA.  One
interesting
finding cited by \citet{Carron2005} is that the absence of crowds results
in overall performance increases. Lastly, as discussed in the introduction, \citet{Lopez}
find evidence that geographic factors like altitude may influence and strengthen home-court
advantage for teams in high altitude regions.

@@

Starting from the top,
Table~\ref{tab:table} summarizes p-values of the nine hypotheses for both
\(z\)-tests and Wilcoxon tests. The p-values are all fairly similar for both
tests giving strong confidence in the conclusions drawn. Also, reported are
the point estimates of the two samples in each comparison.

\jy{add a reference to Bonferroni correction}

@@

There are a few possibilities that might create this increase in two-point shooting
percentage. One could be the possible removal of officiating bias in favor of the 
home team during the bubble. Away teams in the bubble may have benefited from more 
fouls called on drives to the basket. On drives to the basket there can often be a 
lot of contact with no clear foul, these calls are then up to the ref's discretion.
Typically, refs may be more reluctant to blow the whistle against the home team, in
front of their crowd, on this type of play. The NBA only records an attempted shot 
when a foul is called on the shot if the shot is made. So, when away teams are 
fouled, but it is not called because it's often at the ref's discretion to make 
the call, they'll likely miss and be credited the miss. However, in the bubble, if
fouls are called more fairly, the missed shots from uncalled fouls are removed and 
two-point shooting would increase. This wouldn't affect three-point shooting because
those fouls are clearer and less up to the ref's discretion. Another possibility,
without the crowd noise inhibiting their offense, away teams were able to more 
easily run their offensive sets that generate easy two-point looks at the rim. 
Also, away teams being generally more confident without opposing fans present, 
may have been more inclined to attack the basket and get an easy look close to 
the rim. It's hard to say for sure what causes the increase in away two-point shooting,
but these are all possibilities. It may also be a combination of all of these.

>>>>

There are also surveys on the factors contributing to home-court advantage.
\citet{Carron1992} gave four main game location factors for home and away
teams, namely, the crowd factor, which is the impact of fans cheering; learning
factors, which is an advantage from home teams from playing at
a familiar venue; travel factors, the idea that away teams may face
fatigue and jet lag from traveling; and, rule factors, which says that home teams
may benefit from some advantages in rules and officiating. They acknowledge that
these factors would all be removed if games were played at a neutral site, even
if one team was designated as ``home team''. This study was reviewed a decade later
by \citet{Carron2005}. The 2005 review goes over the new findings from studies
about the significance of these four game location factors. Since 1992, they have found
that results on these four factors are mixed. However, there is some evidence
supporting crowd and travel factors impact games across all major sports. There is less
evidence suggesting learning and rule factors impact across the various collegiate 
and professional sports. The NBA is not a league which has rules that
may favor the home team, like batting last in the MLB, but these rule
factors also account for referee bias which may impact the NBA.  One
interesting finding cited by \citet{Carron2005} is that the absence of crowds results
in overall performance increases. Another study by \citet{Price} is able to find 
some evidence of referee bias. They focus on two measures, DTOs and NTOs. DTOs, 
discretionary turnovers, are defined as turnovers always caused by the ref blowing
the whistle while the ball is in play. NTOs, non-discretionary turnovers, are determined 
directly by players with no ref whistle, or the ball going out of bounds. The use
these to test ref bias by checking how variables, like home vs. away team, affect
DTOs relative to NTOs. They found evidence that a home bias does exist. In fact, home 
bias increases both during the playoffs and in games with higher attendance. This is
crucial to the NBA bubble, which consists of playoff games with no attendance. Lastly, 
as discussed in the introduction, \citet{Lopez}
find evidence that geographic factors like altitude may influence and strengthen home-court
advantage for teams in high altitude regions.

@@

Starting from the top,
Table~\ref{tab:table} summarizes p-values of the nine hypotheses for both
\(z\)-tests and Wilcoxon tests. The p-values are all fairly similar for both
tests giving strong confidence in the conclusions drawn. Additionally, the adjusted
p-values, calculated using a Bonferroni correction, and point estimates
for each sample are provided.

@@

There are a few possibilities that might create this increase in two-point shooting
percentage. One could be the possible removal of officiating bias in favor of the 
home team during the bubble. As we saw in \cite{Price}, home bias is normally
stronger in high attendance games. Thus, with no attendance at games, it's reasonable
to say officiating bias was smaller than normal. Away teams in the bubble may have benefited from more 
fouls called on drives to the basket. On drives to the basket there can often be a 
lot of contact with no clear foul, these calls are then up to the ref's discretion.
Typically, refs may be more reluctant to blow the whistle against the home team, in
front of their crowd, on this type of play. The NBA only records an attempted shot 
when a foul is called on the shot if the shot is made. So, when away teams are 
fouled, but it is not called because it's often at the ref's discretion to make 
the call, they'll likely miss and be credited the miss. However, in the bubble, if
fouls are called more fairly, the missed shots from uncalled fouls are removed and 
two-point shooting would increase. This wouldn't affect three-point shooting because
those fouls are clearer and less up to the ref's discretion. Another possibility,
without the crowd noise inhibiting their offense, away teams were able to more 
easily run their offensive sets that generate easy two-point looks at the rim. 
Also, away teams being generally more confident without opposing fans present, 
may have been more inclined to attack the basket and get an easy look close to 
the rim. It's hard to say for sure what causes the increase in away two-point shooting,
but these are all possibilities. It may also be a combination of all of these.
